---
title: "Multivariate Data Analysis"
author: "Juan Pablo Zaldivar && Enric Millán && Joel Solé"
date: "2023-05-11"
output: html_document
---

```{r setup, include=F, warnings=F, echo=F}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE)

YlGnBu = c("#ffffd9", "#edf8b1", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#253494", "#081d58")

source("plots.R")

library(MASS)
library(klaR)
library(dplyr)
library(plotly)
library(ggplot2)
library(qqplotr)
library(caTools)
library(biotools)

# Set the seed for reproducibility
set.seed(123)
```

# Exploratory data analysis


The given dataset consists of measurements of various chemical properties of glass samples, along with their corresponding types. The dataset contains 10 features and 214 instances.

Each row in the dataset corresponds to a specific glass sample and provides the measurements of the aforementioned features, as well as the type of glass it belongs to.

```{r}
glass <- read.csv2("glass.csv", header=TRUE, sep=",")
head(glass)
```

```{r}
# Transform features into numeric
features <- names(glass)[names(glass) != "Type"]
glass[features] <- lapply(glass[features], as.numeric)

# Transform label as categorical
glass$Type <- as.factor(glass$Type)
```

The **type** variable is evidently categorical. It provides information about the different types of glass present in the database. There are six diffrenet levels (for a more detailed explanation of each level see the appendix).

```{r}
unique(glass$Type)
```

```{r}
bar_plot <- plot_ly(data = glass, x = ~Type,
                    color = ~Type, colors = "YlGnBu") %>%
  add_histogram() %>%
  layout(title = "Count of Types in glass Dataset",
         xaxis = list(title = "Type"), yaxis = list(title = "Count"))

bar_plot
```

The dataset is pretty unbalanced. The instances of **types** 1 and 2 constitute more than $67%$ of the glass types.

```{r}
# Compute correlation matrix
corr_matrix <- cor(glass[, 1:9])

# Create a heatmap using Plotly
heatmap <- plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlGnBu"
)

# Add cell values to the heatmap
heatmap <- heatmap %>%
  add_annotations(
    x = rep(colnames(corr_matrix), each = length(colnames(corr_matrix))),
    y = rep(colnames(corr_matrix), length(colnames(corr_matrix))),
    text = round(corr_matrix, 2),
    showarrow = FALSE)

heatmap
```

```{r}
histogram_plots(glass[, -c(10)])
```

Ba and Fe are not very common elements.

```{r}
box_plots(glass[, -c(10)])
```


# Preprocesing

## Data transformation

As seen in the exploratory analysis, the Ba and Fe elements are not very common among all instances. For that reason, we have chosen to binarize those variables, setting them to 1 if the dataframe instance contains Ba and Fe respectively (i.e. the value of the variables is greater than 0) and 0 otherwise.

```{r}
# Transform features into numeric
features <- names(glass)[names(glass) != "Type"]
glass[features] <- lapply(glass[features], as.numeric)

# Transform label as categorical
glass$Type <- as.factor(glass$Type)
```


## Missing values

```{r}
sum(is.na(glass))
```

## Duplicated values

The row number 40 is a duplicate of the previous row.

```{r}
anyDuplicated(glass)
glass <- glass[!duplicated(glass),]
```

```{r}
df_cat <- glass
# Categorize variable 'Ba'
df_cat$Ba <- ifelse(glass$Ba != 0.0, 1, 0)
df_cat$Ba <- as.factor(df_cat$Ba)

# Categorize variable 'Fe'
df_cat$Fe <- ifelse(glass$Fe != 0.0, 1, 0)
df_cat$Fe <- as.factor(df_cat$Fe)

df_cat$Type <- as.factor(df_cat$Type)
```


# PCA

We will apply PCA to the features of the dataset, using the Type variable as supplementary variable

```{r}
library(FactoMineR)

# Assuming your dataset is named "glass"
# Remove the "Type" column from the dataset
numerical_data <- glass[, !names(glass) %in% "Type"]

# Extract the "Type" column as supplementary variable
supplementary_var <- glass$Type

# Apply PCA to the numerical data, specifying the supplementary variable
pca <- PCA(numerical_data, scale = TRUE, quali.sup = supplementary_var)

# Access the results of PCA
summary(pca)      # Summary statistics of the PCA
eig <- get_eigenvalue(pca)     # Eigenvalues
var <- get_pca_var(pca)        # Variables' contributions to dimensions

# Plotting the results (example plot)
plot.PCA(pca, axes = c(1, 2), habillage = glass$Type)

```










.

# Feature description

The features in the dataset are as follows:

RI: Refractive Index (continuous) - It represents the ratio of the speed of light in a vacuum to the speed of light in the glass sample. It is a measure of how much the light is bent as it passes through the glass.

Na: Sodium (continuous) - It denotes the amount of sodium (in weight percent) present in the glass.

Mg: Magnesium (continuous) - It represents the amount of magnesium (in weight percent) present in the glass.

Al: Aluminum (continuous) - It signifies the amount of aluminum (in weight percent) present in the glass.

Si: Silicon (continuous) - It denotes the amount of silicon (in weight percent) present in the glass.

K: Potassium (continuous) - It represents the amount of potassium (in weight percent) present in the glass.

Ca: Calcium (continuous) - It signifies the amount of calcium (in weight percent) present in the glass.

Ba: Barium (continuous) - It represents the amount of barium (in weight percent) present in the glass.

Fe: Iron (continuous) - It denotes the amount of iron (in weight percent) present in the glass.

Type: Type of glass (discrete) - It represents the type of glass sample. There are six distinct types ranging from 1 to 7, excluding the 4th one.

## Type varibale description

Building_windows_float_processed: This level represents glass used for building windows that has undergone a float processing method. Float processing involves floating molten glass on a bed of molten metal to produce a smooth and flat surface.

Building_windows_non_float_processed: This level represents glass used for building windows that has not undergone a float processing method. It may have been processed using alternative methods.

Vehicle_windows_float_processed: This level represents glass used for vehicle windows that has undergone a float processing method. It is specifically processed for automotive applications.

Vehicle_windows_non_float_processed: Although not present in this particular database, this level would represent glass used for vehicle windows that has not undergone a float processing method. Similar to the previous category, it may have been processed using alternative methods.

Containers: This level represents glass used for containers such as bottles or jars.

Tableware: This level represents glass used for tableware items such as plates, bowls, or glasses.

Headlamps: This level represents glass used for headlamp lenses in automotive lighting systems.
