---
title: "MDA_EDA"
author: "Juan Pablo Zaldivar && Enric Millán && Joel Solé"
date: "2023-05-11"
output: html_document
---

```{r setup, include=F, warnings=F, echo=F}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE)

YlGnBu = c("#ffffd9", "#edf8b1", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#253494", "#081d58")

source("plots.R")

library(MASS)
library(klaR)
library(kmed)
library(dplyr)
library(tidyr)
library(HSAUR)
library(httpgd)
library(plotly)
library(ggpubr)
library(ggplot2)
library(qqplotr)
library(caTools)
library(cluster)
library(biotools)
library(gridExtra)
library(factoextra)
library(FactoMineR)

# Set the seed for reproducibility
set.seed(123)
```

# Exploratory data analysis

```{r}
glass <- read.csv2("glass.csv", header=TRUE, sep=",")
head(glass)
```

```{r}
features <- names(glass)[names(glass) != "Type"]
glass[features] <- lapply(glass[features], as.numeric)

glass$Type <- as.factor(glass$Type)
```

```{r}
summary(glass)
```

```{r}
unique(glass$Type)
```

```{r}
bar_plot <- plot_ly(data = glass, x = ~Type,
                    color = ~Type, colors = "YlGnBu") %>%
  add_histogram() %>%
  layout(title = "Count of Types in glass Dataset",
         xaxis = list(title = "Type"), yaxis = list(title = "Count"))

bar_plot
```

```{r}
(sum(glass$Type == 2) + sum(glass$Type == 1))/214
```

```{r}
corr_matrix <- cor(glass[, 1:9])

heatmap <- plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlGnBu"
)

heatmap <- heatmap %>%
  add_annotations(
    x = rep(colnames(corr_matrix), each = length(colnames(corr_matrix))),
    y = rep(colnames(corr_matrix), length(colnames(corr_matrix))),
    text = round(corr_matrix, 2),
    showarrow = FALSE)

heatmap
```

```{r}
histogram_plots(glass[, -c(10)])
```

Ba and Fe are not very common elements.

```{r}
glass_tidy <- gather(data.frame(glass[, 2:9]), key = "Variable", value = "Value")

# Create a boxplot using ggplot2
p <- ggplot(glass_tidy, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(color = "black", outlier.shape = NA) +
  labs(x = "Element", y = "Weight (%)") +
  scale_fill_brewer(palette = "YlGnBu") +
  theme_minimal() +
  theme(legend.position = "none")

p
ggsave("plots/elements_weights.pdf", p, width = 8, height = 6)
```

```{r}
box_plots(glass[, -c(10)])
```

It can be observed that silicon is the primary component of the glass, as it is present on average in more than 70% of the composition.

It is also noteworthy that approximately 90% of the glass composition consists of silicon, sodium, and calcium.

On the other hand, it can be observed that iron and barium are the elements with the least presence.

```{r}
glass_subset <- glass[, c(1:9, 10)]

# Get the list of column names
cols <- names(glass_subset)

# Create a list to store the individual boxplot plots
boxplot_plots <- list()

# Iterate over each column and create a boxplot plot
for (i in 1:9) {
  boxplot_plot <- ggplot(glass_subset, aes_string(x = "Type", y = cols[i])) +
    geom_boxplot(fill = YlGnBu[i], color = "black", size=0.1, outlier.size = 0.6) +
    labs(x = "Type", y = cols[i], title = cols[i]) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(plot.title = element_text(hjust = 0.5, size = 10),  # Center-align the title, adjust font size
          axis.title.x = element_text(size = 8),  # Adjust x-label font size
          axis.title.y = element_text(size = 8))
  
  boxplot_plots[[i]] <- boxplot_plot
}

# Arrange the boxplot plots in a grid
arranged_plots <- grid.arrange(grobs = boxplot_plots, nrow = 3, ncol = 3)

ggsave("plots/feature_boxplots.pdf", arranged_plots, width = 8, height = 8)
```

The analysis of the graphic reveals several notable observations regarding the composition of the glass samples. First, the refractive index falls within a narrow range, ranging from 1.51 to 1.54, indicating consistent optical properties across the samples.

Examining the elemental composition, we notice that Type 6 and Type 7 glasses exhibit higher sodium (Na) percentages compared to other types. Conversely, Type 1, Type 2, and Type 3 glasses display elevated magnesium (Mg) percentages. Additionally, Type 5 and Type 7 glasses showcase higher aluminum (Al) percentages. Interestingly, the silicon (Si) percentage remains relatively consistent across all types, suggesting its integral role as a fundamental component of the glass composition.

Notably, Type 6 glass lacks potassium (K), Barium (Ba) and Iron (Fe) composition entirely. Furthermore, Type 5 and Type 6 glasses exhibit higher calcium (Ca) compositions. Barium (Ba), on the other hand, predominantly appears in Type 7 glasses.

Lastly, iron (Fe) is mainly found in Type 1, 2, and 3 glasses.


# Preprocesing

## Data transformation

As seen in the exploratory analysis, the Ba and Fe elements are not very common among all instances. For that reason, we have chosen to binarize those variables, setting them to 1 if the dataframe instance contains Ba and Fe respectively (i.e. the value of the variables is greater than 0) and 0 otherwise.

```{r}
# # Transform features into numeric
# features <- names(glass)[names(glass) != "Type"]
# glass[features] <- lapply(glass[features], as.numeric)
# 
# # Transform label as categorical
# glass$Type <- as.factor(glass$Type)
```

## Missing values

```{r}
sum(is.na(glass))
```

## Duplicated values

The row number 40 is a duplicate of the previous row.

```{r}
anyDuplicated(glass)
glass <- glass[!duplicated(glass),]
```

```{r}
df_cat <- glass
# Categorize variable 'Ba'
df_cat$Ba <- ifelse(glass$Ba != 0.0, 1, 0)
df_cat$Ba <- as.factor(df_cat$Ba)

# Categorize variable 'Fe'
df_cat$Fe <- ifelse(glass$Fe != 0.0, 1, 0)
df_cat$Fe <- as.factor(df_cat$Fe)

df_cat$Type <- as.factor(df_cat$Type)
```

## Aditional outlier treatment

The detection of outliers was done with the IQR method. For the lower and upper bound of non-outliers point, the value $1.5$ is used as a multiplier to determine the threshold for identifying outliers.

```{r}
head(df_cat)
```

```{r}
num_features <- select_if(df_cat, is.numeric)

outliers <- lapply(num_features, function(feature) {
  q1 <- quantile(feature, 0.25)
  q3 <- quantile(feature, 0.75)
  iqr <- q3 - q1
  lower_bound <- q1 - 3 * iqr 
  upper_bound <- q3 + 3 * iqr
  outlier_rows <- which(feature < lower_bound | feature > upper_bound)
  outlier_rows
})

outlier_indexes <- unique(unlist(outliers))

length(outlier_indexes)

df_cat_clean <- df_cat[-outlier_indexes,]
```

Take into account that the outlier treatment can produce a loss of information, for the outlier samples may contain some valuable information with reference to the glass type. This can potentially introduce a bias in the subsequent analysis. 

Another point to consider is the lack of knowledge in the related field. Although a profound investigation about the glass types and influence of the chemical elements composition was done, we recognize not to be experts in the domain field and this could have a negative effect in the removal of outliers.
