---
title: "MDA_Cluster"
author: "Juan Pablo Zaldivar && Enric Millán && Joel Solé"
date: "2023-05-11"
output: html_document
---

```{r setup, include=F, warnings=F, echo=F}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE)

YlGnBu = c("#ffffd9", "#edf8b1", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#253494", "#081d58")

source("plots.R")

library(MASS)
library(klaR)
library(kmed)
library(dplyr)
library(tidyr)
library(HSAUR)
library(httpgd)
library(plotly)
library(ggpubr)
library(ggplot2)
library(qqplotr)
library(caTools)
library(cluster)
library(biotools)
library(gridExtra)
library(factoextra)
library(FactoMineR)


# Set the seed for reproducibility
set.seed(123)
```

# Exploratory data analysis

```{r}
glass <- read.csv2("glass.csv", header=TRUE, sep=",")
head(glass)
```

```{r}
features <- names(glass)[names(glass) != "Type"]
glass[features] <- lapply(glass[features], as.numeric)

glass$Type <- as.factor(glass$Type)
```

```{r}
summary(glass)
```

```{r}
unique(glass$Type)
```

```{r}
bar_plot <- plot_ly(data = glass, x = ~Type,
                    color = ~Type, colors = "YlGnBu") %>%
  add_histogram() %>%
  layout(title = "Count of Types in glass Dataset",
         xaxis = list(title = "Type"), yaxis = list(title = "Count"))

bar_plot
```

```{r}
(sum(glass$Type == 2) + sum(glass$Type == 1))/214
```

```{r}
corr_matrix <- cor(glass[, 1:9])

heatmap <- plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlGnBu"
)

heatmap <- heatmap %>%
  add_annotations(
    x = rep(colnames(corr_matrix), each = length(colnames(corr_matrix))),
    y = rep(colnames(corr_matrix), length(colnames(corr_matrix))),
    text = round(corr_matrix, 2),
    showarrow = FALSE)

heatmap
```

```{r}
histogram_plots(glass[, -c(10)])
```

```{r}
glass_tidy <- gather(data.frame(glass[, 2:9]), key = "Variable", value = "Value")

# Create a boxplot using ggplot2
p <- ggplot(glass_tidy, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(color = "black", outlier.shape = NA) +
  labs(x = "Element", y = "Weight (%)") +
  scale_fill_brewer(palette = "YlGnBu") +
  theme_minimal() +
  theme(legend.position = "none")

p
ggsave("plots/elements_weights.pdf", p, width = 8, height = 6)
```

```{r}
box_plots(glass[, -c(10)])
```

```{r}
glass_subset <- glass[, c(1:9, 10)]

# Get the list of column names
cols <- names(glass_subset)

# Create a list to store the individual boxplot plots
boxplot_plots <- list()

# Iterate over each column and create a boxplot plot
for (i in 1:9) {
  boxplot_plot <- ggplot(glass_subset, aes_string(x = "Type", y = cols[i])) +
    geom_boxplot(fill = YlGnBu[i], color = "black", size=0.1, outlier.size = 0.6) +
    labs(x = "Type", y = cols[i], title = cols[i]) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(plot.title = element_text(hjust = 0.5, size = 10),  # Center-align the title, adjust font size
          axis.title.x = element_text(size = 8),  # Adjust x-label font size
          axis.title.y = element_text(size = 8))
  
  boxplot_plots[[i]] <- boxplot_plot
}

# Arrange the boxplot plots in a grid
arranged_plots <- grid.arrange(grobs = boxplot_plots, nrow = 3, ncol = 3)

ggsave("plots/feature_boxplots.pdf", arranged_plots, width = 8, height = 8)
```


# Preprocesing

## Data transformation

```{r}
# # Transform features into numeric
# features <- names(glass)[names(glass) != "Type"]
# glass[features] <- lapply(glass[features], as.numeric)
# 
# # Transform label as categorical
# glass$Type <- as.factor(glass$Type)
```

## Missing values

```{r}
sum(is.na(glass))
```

## Duplicated values

```{r}
anyDuplicated(glass)
glass <- glass[!duplicated(glass),]
```

```{r}
df_cat <- glass
# Categorize variable 'Ba'
df_cat$Ba <- ifelse(glass$Ba != 0.0, 1, 0)
df_cat$Ba <- as.factor(df_cat$Ba)

# Categorize variable 'Fe'
df_cat$Fe <- ifelse(glass$Fe != 0.0, 1, 0)
df_cat$Fe <- as.factor(df_cat$Fe)

df_cat$Type <- as.factor(df_cat$Type)
```

## Aditional outlier treatment

```{r}
head(df_cat)
```

```{r}
num_features <- select_if(df_cat, is.numeric)

outliers <- lapply(num_features, function(feature) {
  q1 <- quantile(feature, 0.25)
  q3 <- quantile(feature, 0.75)
  iqr <- q3 - q1
  lower_bound <- q1 - 3 * iqr 
  upper_bound <- q3 + 3 * iqr
  outlier_rows <- which(feature < lower_bound | feature > upper_bound)
  outlier_rows
})

outlier_indexes <- unique(unlist(outliers))

length(outlier_indexes)

df_cat_clean <- df_cat[-outlier_indexes,]
```


# Clusters

```{r}
df_num  <- subset(df_cat, select = -c(Ba, Fe, Type))
summary(df_num)
var(df_num$RI)
```

Since all variables that we are gonna use for clustering, except for RI(Reflective Index), have the same units it is not advisable to scale them because that may hide some variance and inner groups. Moreover, reflective Index has a really low variance so it is probably not very important to make clusters.

```{r}
d <- dist(df_num, method = "euclidean") # distance matrix
fit <- hclust(d, method="single")
plot(fit, main = "Dendrogram of Single Linkage", labels = FALSE) 
fit1 <- hclust(d, method="complete") 
plot(fit1, main = "Dendrogram of complete Linkage", labels = FALSE) 
fit2 <- hclust(d, method="average") 
plot(fit2, main = "Dendrogram of Average Linkage", labels = FALSE)
fit3 <- hclust(d, method="ward.D2") 
plot(fit3, main = "Dendrogram of Ward Method", labels = FALSE) 
fit4 <- hclust(d, method="centroid") 
plot(fit4, main = "Dendrogram of Centroid Method", labels = FALSE) # Dendogram 
```

Ward method seems to give the most reasonable clustering of the dataset. With this method, two distinct groups can be clearly observed when considering a height of 20. Additionally, three groups can be considered if we use a lower height, like 15. This clear differentiation, which was not seen with previous methods, is due to the robustness of this method in the presence of outliers and a lot of samples.

## Highlighting groups

The Ward method is the one that provides us with the greatest clarity regarding the groups that can be formed in the dataset. Therefore, we will use it as the basis for our analysis in hierarchical clustering.

```{r}
plot(fit3,main="Dendrogram Ward Method Linkage", labels=FALSE)
groups <- cutree(fit3, k=2 )# c
rect.hclust(fit3, k=2, border=YlGnBu[2])

groups <- cutree(fit3, k=3 )# c
rect.hclust(fit3, k=3, border=YlGnBu[4])

groups <- cutree(fit3, k=4 )# c
rect.hclust(fit3, k=4, border=YlGnBu[6])

groups <- cutree(fit3, k=5 )# c
rect.hclust(fit3, k=5, border=YlGnBu[8])

groups <- cutree(fit3, k=6 )# c
rect.hclust(fit3, k=6, border=YlGnBu[10])
```

Looks like the most reasonable option is 3 groups, since adding more groups creates some of them with very few samples, which may be due to the presence of some outliers. Nevertheless, let's confirm this hypothesis with the Elbow Graph.

### Elbow Graph

```{r}
aux<-c()
for (i in 1:dim(df_num)[2]) {
  k <- kmeans(df_num, centers = i, nstart = 25)
  aux[i] <- k$tot.withinss
}
plot(aux, xlab="Number of Clusters", ylab="TWSS", type="l", main="TWSS vs. number of clusters", col=YlGnBu[6], lwd=3)

fviz_nbclust(df_num, kmeans, method = "silhouette")
fviz_nbclust(df_num, kmeans, method = "gap_stat")
```

As wee can see, with different criteria: "wss", "silhouette" and "gap_stat", the result seems to be the same, the optimal number of clusters is 3 as we theorized previously. 

### Silhouette Index

```{r}
res <- fastkmed(d, 6)
silhouette <- sil(d, res$medoid, res$cluster)

silhouette$plot
```

Silhouette index confirms that the most adequate number of clusters is 3.

```{r}
cluster = cutree(fit3,3)

p <- ggplot(data = df_num, aes(x = cluster, fill = glass$Type)) +
  geom_bar() +
  theme_bw() + labs(title="", fill = "Type") + scale_fill_brewer(palette = 'YlGnBu')

plot(p)
```

Keep in mind this distribution of types of glass around clusters because we will see that hclust cut in 3 groups coincides with kmeans for 3 groups.

## Kmeans

```{r}
k3 <- kmeans(d, centers = 3, nstart = 25)
str(k3)
names(k3)

aggregate(df_num,by=list(k3$cluster),FUN=mean)
df_num$cluster<-as.numeric(k3$cluster)
```

We observe that variables like _RI_, _Na_, _Al_ and _Si_ don't play an important role for differentiating the groups. On the other hand, variables such as _Mg_ outstand with a practically null value in group 1, a  moderate value in group 2, and a big value in comparison in group 3, being a variable that seems to be important for the clustering of the groups. Another variable important to separate groups might be _K_, which seems to separate group 1 from the other two. Something similar happens with _Ca_, which has a pretty high value for group 1 and similar values for groups 2 and 3.

```{r}
k3$withinss
k3$totss
k3$tot.withinss
k3$betweenss + k3$tot.withinss # BSS + Wss
```

### Kmeans Plots

```{r}
fviz_cluster(k3, data = df_num,
             palette = c(YlGnBu[2], YlGnBu[6], YlGnBu[9]), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )
```

```{r}
names(k3)
clusplot(df_num, k3$cluster, color=TRUE, shade=TRUE,
         labels=2, lines=0)

c3<-clara(df_num,3)
names(c3)

clusplot(df_num, c3$cluster, color=TRUE, shade=TRUE,
         labels=2, lines=0)
```

```{r}
fviz_cluster(k3, data = df_num,
             palette = c("#2E9FDF", "#00AFBB", "#E7B800"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw(),
             )
```

The clusters formed with kmeans seem to separate more properly the samples than the ones generated with 'clara' method. 

```{r}
p <- ggplot(data = df_num, aes(x = k3$cluster, fill = glass$Type)) +
  geom_bar() +
  theme_bw()

print(p)
```

Let's see if the clustering separates with clarity some of the different types of glass. Glass type 2, which goes for non-processed building windows, seems to be the most common in all groups, but for instance, group three gathers all of glass type 1 and 3, which are building and vehicle windows but most importantly both processed. Meanwhile group 2 contains most of types 5, 6 and 7, which are the types belonging to containers, tableware and headlamps, which happen to be the types less similar to vehicle or building windows. Cluster 1 seems to be the more diffuse, but we already saw in the cluster plot that is the one with less samples and is has a considerable intersection with group 2, so it makes sense that is the one that gives less information.

