---
title: "MDA_PCA"
author: "Juan Pablo Zaldivar && Enric Millán && Joel Solé"
date: "2023-05-11"
output: html_document
---

```{r setup, include=F, warnings=F, echo=F}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE)

YlGnBu = c("#ffffd9", "#edf8b1", "#c7e9b4", "#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#253494", "#081d58")

source("plots.R")

library(MASS)
library(klaR)
library(kmed)
library(dplyr)
library(tidyr)
library(HSAUR)
library(httpgd)
library(plotly)
library(ggpubr)
library(ggplot2)
library(qqplotr)
library(caTools)
library(cluster)
library(biotools)
library(gridExtra)
library(factoextra)
library(FactoMineR)

# Set the seed for reproducibility
set.seed(123)
```

# Exploratory data analysis

```{r}
glass <- read.csv2("glass.csv", header=TRUE, sep=",")
head(glass)
```

```{r}
features <- names(glass)[names(glass) != "Type"]
glass[features] <- lapply(glass[features], as.numeric)

glass$Type <- as.factor(glass$Type)
```

```{r}
summary(glass)
```

```{r}
unique(glass$Type)
```

```{r}
bar_plot <- plot_ly(data = glass, x = ~Type,
                    color = ~Type, colors = "YlGnBu") %>%
  add_histogram() %>%
  layout(title = "Count of Types in glass Dataset",
         xaxis = list(title = "Type"), yaxis = list(title = "Count"))

bar_plot
```

```{r}
(sum(glass$Type == 2) + sum(glass$Type == 1))/214
```

```{r}
corr_matrix <- cor(glass[, 1:9])

heatmap <- plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlGnBu"
)

heatmap <- heatmap %>%
  add_annotations(
    x = rep(colnames(corr_matrix), each = length(colnames(corr_matrix))),
    y = rep(colnames(corr_matrix), length(colnames(corr_matrix))),
    text = round(corr_matrix, 2),
    showarrow = FALSE)

heatmap
```

```{r}
histogram_plots(glass[, -c(10)])
```

```{r}
glass_tidy <- gather(data.frame(glass[, 2:9]), key = "Variable", value = "Value")

# Create a boxplot using ggplot2
p <- ggplot(glass_tidy, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(color = "black", outlier.shape = NA) +
  labs(x = "Element", y = "Weight (%)") +
  scale_fill_brewer(palette = "YlGnBu") +
  theme_minimal() +
  theme(legend.position = "none")

p
ggsave("plots/elements_weights.pdf", p, width = 8, height = 6)
```

```{r}
box_plots(glass[, -c(10)])
```

```{r}
glass_subset <- glass[, c(1:9, 10)]

# Get the list of column names
cols <- names(glass_subset)

# Create a list to store the individual boxplot plots
boxplot_plots <- list()

# Iterate over each column and create a boxplot plot
for (i in 1:9) {
  boxplot_plot <- ggplot(glass_subset, aes_string(x = "Type", y = cols[i])) +
    geom_boxplot(fill = YlGnBu[i], color = "black", size=0.1, outlier.size = 0.6) +
    labs(x = "Type", y = cols[i], title = cols[i]) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(plot.title = element_text(hjust = 0.5, size = 10),  # Center-align the title, adjust font size
          axis.title.x = element_text(size = 8),  # Adjust x-label font size
          axis.title.y = element_text(size = 8))
  
  boxplot_plots[[i]] <- boxplot_plot
}

# Arrange the boxplot plots in a grid
arranged_plots <- grid.arrange(grobs = boxplot_plots, nrow = 3, ncol = 3)

ggsave("plots/feature_boxplots.pdf", arranged_plots, width = 8, height = 8)
```


# Preprocesing

## Data transformation

```{r}
# # Transform features into numeric
# features <- names(glass)[names(glass) != "Type"]
# glass[features] <- lapply(glass[features], as.numeric)
# 
# # Transform label as categorical
# glass$Type <- as.factor(glass$Type)
```

## Missing values

```{r}
sum(is.na(glass))
```

## Duplicated values

```{r}
anyDuplicated(glass)
glass <- glass[!duplicated(glass),]
```

```{r}
df_cat <- glass
# Categorize variable 'Ba'
df_cat$Ba <- ifelse(glass$Ba != 0.0, 1, 0)
df_cat$Ba <- as.factor(df_cat$Ba)

# Categorize variable 'Fe'
df_cat$Fe <- ifelse(glass$Fe != 0.0, 1, 0)
df_cat$Fe <- as.factor(df_cat$Fe)

df_cat$Type <- as.factor(df_cat$Type)
```

## Aditional outlier treatment

```{r}
head(df_cat)
```

```{r}
num_features <- select_if(df_cat, is.numeric)

outliers <- lapply(num_features, function(feature) {
  q1 <- quantile(feature, 0.25)
  q3 <- quantile(feature, 0.75)
  iqr <- q3 - q1
  lower_bound <- q1 - 3 * iqr 
  upper_bound <- q3 + 3 * iqr
  outlier_rows <- which(feature < lower_bound | feature > upper_bound)
  outlier_rows
})

outlier_indexes <- unique(unlist(outliers))

length(outlier_indexes)

df_cat_clean <- df_cat[-outlier_indexes,]
```


# PCA

In order to properly apply Principal Component Analysis (PCA), our data needs to meet certain assumptions. First, we require our data to follow a normal distribution. We observe that in our case, this assumption is not met for some variables, particularly for Ba and Fe. As mentioned earlier, these variables have a large number of zero values. Furthermore, we need a certain level of correlation among variables. However, the variable Fe does not exhibit significant correlations with the rest of the variables. Therefore, we have decided to exclude the Ba and Fe variables from this analysis.

Additionally, it is essential for our data to be free from outliers. Hence, before applying PCA, we will remove outliers using the Interquartile Range (IQR) method. We used a larger margin in order to avoid removing too many observations.

<!-- Ya hay un tratamiento de outliers en el preprocessing, porque no usas ese? @JOEl? -->

```{r}
df_pca <- df_cat[,-c(8, 9)]

num_features <- select_if(df_pca, is.numeric)

# Detect outliers using the IQR method
outliers <- lapply(num_features, function(feature) {
  q1 <- quantile(feature, 0.25)
  q3 <- quantile(feature, 0.75)
  iqr <- q3 - q1
  lower_bound <- q1 - 3 * iqr 
  upper_bound <- q3 + 3 * iqr
  outlier_rows <- which(feature < lower_bound | feature > upper_bound)
  outlier_rows
})

# Combine the outlier indexes
outlier_indexes <- unique(unlist(outliers))

length(outlier_indexes)

# Drop the outlier rows from the "glass" dataframe
df_pca <- df_pca[-outlier_indexes,]

# df_pca == df_cat_clean[,-c(8,9)]
# same df
```

We will apply PCA to the features of the dataset, using the Type variable as supplementary variable and RI as supplementary variable.

```{r}
 pca <- PCA(df_pca, scale = TRUE, quali.sup = 8, quanti.sup = 1, graph=F)
```


## Component extraction

```{r}
pca$eig
```

We applied PCA to the cleaned data and for the analysis of the resulting components, we will focus on the first three components. These three components were chosen because they have eigenvalues above the mean. By utilizing the first three components, we are able to explain 85% of the variance in the data.

```{r}
variance_percent <- pca$eig[,2]

# Calculate the cumulative percentage of variance
cumulative_var <- cumsum(variance_percent)

data <- data.frame(Components = seq_along(variance_percent),
                   VariancePercent = variance_percent,
                   CumulativeVar = cumulative_var)

# Plotting with ggplot2
p <- ggplot(data, aes(x = Components)) +
  geom_bar(aes(y = VariancePercent), stat = "identity", fill = YlGnBu[6], width = 0.75) +
  geom_line(aes(y = CumulativeVar), color = YlGnBu[4], group = 1, lwd = 1) +
  geom_point(aes(y = CumulativeVar), color = YlGnBu[5], size = 3) +
  labs(x = "Number of Components", y = "Percentage of Variance",
       title = "Cumulative Percentage of Variance Explained") +
  geom_hline(yintercept = mean(variance_percent), linetype = "dashed", color = YlGnBu[9], lwd=1) +
  ylim(0, 100) +
  theme_minimal()+
  scale_x_continuous(breaks = seq_along(variance_percent))

p
ggsave("plots/var_cumvar.pdf", p, width = 8, height = 5)
```

Since the data is scaled, the mean eigenvalue is 1. Therefore, we will extract the first 4 principal components since they are the only components to have an eigenvalue above 1.

## Correlation between dimension

```{r}
pca$var$cor
```

```{r}
p1 <- plot.PCA(pca, axes=c(1,2), choix="var")
p2 <- plot.PCA(pca, axes=c(2,3), choix="var")
p3 <- plot.PCA(pca, axes=c(1,3), choix="var")
p1; p2; p3

ggsave("plots/pca_dim12.pdf", p1, width = 5, height = 5)
ggsave("plots/pca_dim23.pdf", p2, width = 5, height = 5)
ggsave("plots/pca_dim13.pdf", p3, width = 5, height = 5)
```

## Individual plots

```{r}
YlGnBu_modified = c( "#edf8b1", "#7fcdbb", "#41b6c4", "#1d91c0", "#253494", "#081d58")

plot_pca <- function(pca, df_pca, dim1, dim2, save = "") {
  df_temp <- data.frame(pca$ind$coord[, c(dim1, dim2)], df_pca$Type)
  colnames(df_temp) <- c("x1", "x2", "Type")

  plot <- ggplot(df_temp, aes(x = x1, y = x2, color = Type, shape = Type)) +
    geom_point(size = 2) +
    labs(x = paste("Component", dim1), y = paste("Component", dim2), color = "Type") +
    scale_shape_manual(values = c(0, 1, 2, 15, 16, 17)) +
    theme_minimal() +
    scale_color_manual(values = YlGnBu_modified) +
    guides(color = guide_legend(override.aes = list(shape = c(0, 1, 2, 15, 16, 17))), shape = "none")
  
  if (save != "") {
    ggsave(save, plot,  width = 8, height = 6)
  }
  print(plot)
}

plot_pca(pca, df_pca, 1, 2, save="plots/pca_scatter_ind12.pdf")
plot_pca(pca, df_pca, 2, 3, save="plots/pca_scatter_ind23.pdf")
plot_pca(pca, df_pca, 1, 3, save="plots/pca_scatter_ind13.pdf")
```
